Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
whack a bit till stuff fits into 8 pages

An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator.  The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers.\footnote{The
chain of \texttt{if}, \texttt{elif}, ... instructions that check the various
opcodes is transformed into a \texttt{switch} statement by one of PyPy's
optimizations. Python does not have a \texttt{switch} statement}
A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

Machine code production is done via a well-defined interface to an assembler
backend. This makes it possible to easily port the tracing JIT to various
architectures (including, we hope, to virtual machines such as the JVM where
our backend could generate JVM bytecode at runtime). At the moment the only
implemented backend is a 32-bit Intel-x86 backend.

@@

In this section we evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1 GB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include running the tracer and producing machine code.

@@

\textbf{Benchmark 3:} The tracing JIT is enabled and hints as in Figure
\ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
so that it corresponds to the loop in the square function. However, constant folding of green
variables is disabled, therefore the resulting machine code corresponds to the
trace in Figure \ref{fig:trace-no-green-folding}. This alone brings an
improvement over the previous case, but is still slower than pure
interpretation.

>>>>

An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator.  The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers.\footnote{The
chain of \texttt{if}, \texttt{elif}, ... instructions checking the various
opcodes is turned into a \texttt{switch} statement by one of PyPy's
optimizations. Python does not have a \texttt{switch} statement.}
A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

Machine code production is done via a well-defined interface to an assembler
backend. This allows easy porting of the tracing JIT to various
architectures (including, we hope, to virtual machines such as the JVM where
our backend could generate JVM bytecode at runtime). At the moment the only
implemented backend is a 32-bit Intel-x86 backend.

@@

In this section we evaluate the work done so far by looking at some
benchmarks. Since the work is not finished, these can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1 GB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include running the tracer and producing machine code.

@@

\textbf{Benchmark 3:} The tracing JIT is enabled and hints as in Figure
\ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
so that it corresponds to the loop in the square function. Constant folding of green
variables is disabled, therefore the resulting machine code corresponds to the
trace in Figure \ref{fig:trace-no-green-folding}. This alone brings an
improvement over the previous case, but is still slower than pure
interpretation.
